### 2021-07-11 docs/manuscript/manuscript.tex
Reformat and add citations

Our understanding of how GRNs process information has increased thanks to the
concept and theory of network motifs, defined as patterns of
interconnections occurring in networks at numbers significantly higher
than those in randomized networks~\cite{Milo2002}. This novel approach, along
with mathematical modeling and computational systems biology methods, is a
powerful tool to study GRNs~\cite{kitano_computational_2002}, and
experimental studies have validated the predictions of studies using these
methods~\cite{Kalir2005,Kalir2004,Mangan2003a,ODonnell2005}.
Some have even used synthetic GRNs to produce artificial cell
differentiation~\cite{Basu2005}, circadian gene
expression~\cite{atkinson_development_2003}, counting
devices~\cite{friedland_synthetic_2009} and systems that respond to light.
Moreover, the design and implementation of GRNs through synthetic biology is
emerging as a promising tool to study biological phenomena as pattern
formation~\cite{santos-moreno_using_2019} and as a novel therapeutic tool with
interesting biomedical applications~\cite{karlsson_therapeutic_2012,
higashikuni_advancing_2017,abil_synthetic_2015,healy_genetic_2019,
kitada_programming_2018}.

@@

Although GRNs with positional information processing capacities have been
extensively analyzed, most studies have so far emphasized particular
regulatory systems that respond to a single input only. These studies have
typically focused on either regulatory systems observed in
nature~\cite{vonDassow2000,Reinitz1995,Jaeger2004}, synthetic
implementations~\cite{Basu2005,Schaerli2018,Elowitz2000,barbier2020} with
predefined topologies, or have deliberately constrained the study of GRNs to just
a handful of alternative designs~\cite{munteanu_2014,Schaerli2014}. However,
previous work has demonstrated the necessity to expand computational analysis
of pattern forming GRNs to multi-input settings. For instance, it has been shown
that the neural subtypes specification system in the vertebrate neural tube involves
a  two-input network in which Sonic Hedgehog acts as a morphogen and the Olig2
and Nkx2.2 genes can act as the receiver nodes~\cite{dessaud_dynamic_2010,
balaskas2012, exelby_2021}.

In this study, we apply a Markov chain Monte Carlo (MCMC)-like algorithm to
probe the design space of three-node GRNs looking for topologies capable of
translating a morphogen gradient  (input signal) into a striped pattern of gene
expression (phenotype). Importantly, unlike most previous computational
studies, here we allowed any node (e.g., gene) in a GRN to perceive the input
signal in varying ways and selected the best-performing GRNs based on a
fitness criterion used to assess the quality of the phenotype to
a prescribed optimal pattern. Based on this computational strategy, we uncovered
a great variety of distinct classes of network topologies that
tend to form a complex interconnected meta-graph that could be easily
traversed throughout evolution via single changes in the wiring of the
different GRNs.

@@

In our model the morphogen could interact with any of the three genes on the
GRN, but could not be affected by them. The morphogenetic field was defined by
an unidimensional array of 30 isogenic cells exposed to the morphogen
concentration gradient (Fig~\ref{fig:model}C). The initial concentration of
each gene product in all cells was set to 1 in all simulations.\\

@@

In order to estimate the robustness of topology 9 to changes in parameters that
control morphogen gradient and changes in initial concentrations of gene
products, we generated 100 fitness values for each one of the 19 GRNs
belonging to topology 9; in each of these 100 assays we chose $A_0$, $h$ or
initial concentrations randomly from a Normal Distribution with $\mu = 1$ for
$A_0$, $\mu = 0.4$ for $h$ and $\mu = 1$ for initial concentrations and a
Coefficient of Variation of 30\% in each case.

>>>>

Our understanding of how GRNs process information has increased thanks to the
concept and theory of network motifs, defined as patterns of
interconnections occurring in networks at numbers significantly higher
than those in randomized networks~\cite{Milo2002}. This novel approach, along
with mathematical modeling and computational systems biology methods, is a
powerful tool to study GRNs~\cite{kitano_computational_2002}, and
experimental studies have validated the predictions of studies using these
methods~\cite{Kalir2005,Kalir2004,Mangan2003a,ODonnell2005}.
Some have even used synthetic GRNs to produce artificial cell
differentiation~\cite{Basu2005}, circadian gene
expression~\cite{atkinson_development_2003}, counting
devices~\cite{friedland_synthetic_2009} and systems that respond to
light~\cite{gardner2012,levskaya2005}.
Moreover, the design and implementation of GRNs through synthetic biology is
emerging as a promising tool to study biological phenomena as pattern
formation~\cite{santos-moreno_using_2019} and as a novel therapeutic tool with
interesting biomedical applications~\cite{karlsson_therapeutic_2012,
higashikuni_advancing_2017,abil_synthetic_2015,healy_genetic_2019,
kitada_programming_2018}.

@@

Although GRNs with positional information processing capacities have been
extensively analyzed, most studies have so far emphasized particular
regulatory systems that respond to a single input only. These studies have
typically focused on either regulatory systems observed in
nature~\cite{vonDassow2000,Reinitz1995,Jaeger2004}, synthetic
implementations with predefined topologies~\cite{Basu2005, Schaerli2018,
Elowitz2000, barbier2020}, or have deliberately constrained the study of GRNs to just
a handful of alternative designs~\cite{munteanu_2014,Schaerli2014}. However,
previous work has demonstrated the necessity to expand computational analysis
of pattern forming GRNs to multi-input settings. For instance, it has been shown
that the neural subtypes specification system in the vertebrate neural tube involves
a  two-input network in which Sonic Hedgehog acts as a morphogen and the Olig2
and Nkx2.2 genes can act as the receiver nodes~\cite{dessaud_dynamic_2010,
balaskas2012, exelby_2021}.

In this study, we apply a Markov chain Monte Carlo (MCMC)-like algorithm to
probe the design space of three-node GRNs looking for topologies capable of
translating a morphogen gradient  (input signal) into a striped pattern of gene
expression (phenotype). Importantly, unlike most previous computational
studies, here we allowed any node (e.g., gene) in a GRN to perceive the input
signal in varying ways and selected the best-performing GRNs based on a
fitness criterion used to assess the quality of the phenotype with respect to
a prescribed optimal pattern. Based on this computational strategy, we uncovered
a great variety of distinct classes of network topologies that
tend to form a complex interconnected meta-graph that could be easily
traversed throughout evolution via single changes in the wiring of the
different GRNs.

@@

In our model the morphogen could interact with any of the three genes on the
GRN, but could not be affected by them. The morphogenetic field was defined by
an unidimensional array of 30 isogenic cells exposed to the morphogen
concentration gradient (Fig~\ref{fig:model}C). The initial concentration of
each gene product in all cells was set to 0.1 in all simulations.\\

@@

In order to estimate the robustness of topology 9 to changes in parameters that
control morphogen gradient and changes in initial concentrations of gene
products, we generated 100 fitness values for each one of the 19 GRNs
belonging to topology 9; in each of these 100 assays we chose $A_0$, $h$ or
initial concentrations randomly from a Normal Distribution with $\mu = 1$ for
$A_0$, $\mu = 0.4$ for $h$ and $\mu = 0.1$ for initial concentrations and a
Coefficient of Variation of 30\% in each case.
